Batson Challenges Using Historical Strike Data

### 2021-08-03 analysis/paper_draft.Rmd
updated intro text

---
title: "Batson Challenges Using Historical Data"
date: "`r Sys.Date()`"
bibliography: references.bib
author:
  - name: "Sachin S. Pandya"
    affiliation: "University of Connecticut"
    orcid_id: 0000-0001-7387-1307
  - name: "Xiaomeng Li"
    affiliation: "University of Connecticut"
  - name: "Timothy E. Moore"
    affiliation: "University of Connecticut"
    orcid_id: 0000-0002-9576-0517
output: distill::distill_article
---

@@

This paper shows how to use historical jury-selection data to estimate how likely an attorney has used peremptory challenges to strike potential jurors because of juror race or gender in violation of _Batson v. Kentucky_ (1986), and its progeny.

In the US, once a group of individuals selected for jury-service appear in court as scheduled, this group is questioned by the parties' attorneys and the trial judge.  During this "voir dire" process, a prospective juror, if not selected for the jury or excused by the trial judge for cause, may be dismissed if a party's attorney uses a peremptory challenge against them.

@@

By law, evidence supporting a _Batson_ challenge may include not only the circumstances of the jury selection itself, but also data on the pattern of peremptory challenges in past cases. For example, in a _Batson_ challenge against a prosecutor in a criminal case, such historical data would be relevant to whether (1) to conclude that the defendant has a prima facie case; (2) to believe a prosecutor from that office who has offered a race-neutral explanation for striking, say, an African-American juror in that case; and (3) to ultimately accept or reject a Batson challenge in that case [e.g., Miller-El v. Dretke, 545 U.S. 231, 253 (2005); Riley v. Taylor, 277 F.3d 261, 280-81, 283 (3d Cir. 2001); Cochran v. Herring, 43 F.3d 1404, 1412 (11th Cir. 1995)]. The US Supreme Court recently identified the kinds of evidence to support _Batson_ challenge as including "relevant history of the State's peremptory strikes in past cases" [e.g, Flowers v. Mississippi, 139 S. Ct. 2228, 2243 (2019)].

Prior work on statistical methods for _Batson_ challenges, however, has largely focused on strikes in a single case [e.g. @Barrett2007; @Gastwirth2016; @Kadane2018; @Kadane2018a]. When considering strike data in multiple trials from the same State or judicial district, @Gastwirth2013 criticized the use of a paired t-test to compare the differences in strike rates of Black and non-Black venire members in the aggregate. Rather, both they and @Gastwirth2014 argued for applying the Cochran–Mantel–Haenszel test by taking the difference of the sum of the observed numbers of Black venire members challenged in each trial and the corresponding sum of the expected numbers of Blacks challenged, and then comparing it to the standard deviation of the difference.

Part of the issue is that strike data from past cases is hard to collect. Jury-selection records are often not readily available and largely on paper, not electronic or machine-readable formats [@Grosso2017; @Wright2018], and otherwise require painstaking efforts to collect [e.g., @Craft2018].  As a result, even where there, for any given attorney, available strike data on past trials, that data may be missing strikes from trials not at random. Possible reasons include limited access the relevant records (e.g., jury-selection records kept by a jury administrator, not in a case file) and the expense of obtaining copies.  Some jurisdictions do not require recording relevant or detailed enough information about jurors who were or could have been struck, such as their race or gender. As a result, lawyers and researchers must expend time and effort to access other data sources (e.g., voter registration rolls) with that information. Finally, regardless of record-keeping policy, judges and court staff vary in how often they fully complete or maintain the required information.

@@

Here, we extend the Bayesian approach to _Batson_ challenges to assess whether one can credibly infer illegal bias based solely on an attorney's pattern of peremptory challenges using both present strike data and historical data that attorney's strikes in past cases.

In short, our workflow is this: First, specify a model of the peremptory-strike process that best approximates the peremptory-strike process in the jurisdiction of interest. Second, assign a prior distribution to the bias parameter $b$ that reflects the legal requirement that the _Batson_ challenger bear the burden of proof. Here, we assume a normal prior for $b$ with mean 0 and standard deviation 2. Third, using the strike data in past trials generated by the same peremptory-strike process, estimate a posterior distribution for $b$.  Finally, using data on the target attorney's strikes in the present case, use that posterior distribution as the prior for $b$ and update.

In so doing, we make two main contributions. First, we use the power prior to control how much the strike history in past cases affects the posterior distribution of the bias parameter $b$. The power prior is a way to create informative prior distributions based on historical data so as to control how much historical data affects the posterior distribution of the unobserved variables of interest. The key insight is to raise the likelihood function based on the historical data to a power parameter [@Ibrahim2000; @Ibrahim2015].

@@

Following @Kadane2018 and @Barrett2007, we model a struck-jury peremptory-challenge process as follows.

For any given jury selection $i$, let $j$ denote a peremptory strike used, and let $\delta_{ij}$ denote whether or not a party used that strike on a person who belongs to a "cognizable class". If "race" is the bias type of interest, the cognizable class is racial minority jurors ($\delta_{ij} =1$, 0 for White jurors).  If "gender" is the bias of interest, the cognizable class is female jurors ($\delta_{ij} = 1$, 0 for male jurors).

In turn, let $c_{ij}$ denote the number of cognizable class members subject to strike; and $m_{ij}$ denote the number of cognizable class non-members subject to strike, such that $c_{ij}+m_{ij}$ is the total number of jurors potentially subject to strike. If there is no bias, the probability is $\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking a cognizable class member, and $1-\frac{c_{ij}}{c_{ij}+m_{ij}}$.

By adding one parameter $w$, we can measure the bias by different values of $w$ by defining the probability of a cognizable class member being struck to be $\frac{wc_{ij}}{wc_{ij}+m_{ij}}$. As a result, however, the weight of the non-cognizable class is the reciprocal of the weight of cognizable class. Using $b = log(w)$ avoids this issue. Accordingly, for any given value of the bias parameter $b$, the probability of strike of a member from either class, or $Pr(\delta_{ij})$, is shown in equation \@ref(eq:model1)

@@

We also generated historical strike data of three sizes (same, double, and triple the size of current data) as well as seven different values of the bias parameter for generating the historical data ("historical" $b = {-3,-2,-1,0,1,2,3}$).

We let historical $b$ differ from current $b$ (our parameter of interest) for two reasons. First, it allows us to simulate noise associated with historical strike data that is incomplete not at random. Second, it could also simulate a real difference in the bias parameter that depends on a feature that the current trial has but that _all_ the past trials in the historical data do not. To illustrate, assume a prosecutor with two true bias values: $b > 0$ if the criminal defendant is Black, $b = 0$ if not. If our historical data contains only strike data in cases with non-Black defendants, then that prosecutor's historical $b$ will differ from the current case with a Black defendant.

By using the power prior, we can control the weight that we put on historical data by modifying the weight parameter $a$. If $a = 1$, the historical strike data is equally weighted with the present data on strikes. If $a < 1$, the historical data matters proportionally less. As discussed above, if some of the historical data is missing not at random, we may wish to assign $a < 1$ to evaluate how sensitive the posterior of the bias parameter to the historical data on strikes. Accordingly, in the simulation study, we tried different values of the weight parameter $a = (0.1,0.3,0.5,0.7,1)$ to show how that parameter affected modeling accuracy.

@@

In turn, we set three credible interval level (80\%, 90\% and 95\%) and recorded the coverage rates, i.e., the proportion of times those intervals actually contained the true value of the (current) bias parameter $b$. For instance, a 90\% coverage rate means that in 900 of 1000 model fits, a given credible interval contained what we fixed as the value of current $b$. Put another way, if we set $b>0$, we calculate the proportion of the 90\% credible intervals that lie on the right of 0 among the 1000 model fits.

>>>>

---
title: "Batson Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
bibliography: references.bib
author:
  - name: "Sachin S. Pandya"
    affiliation: "University of Connecticut"
    orcid_id: 0000-0001-7387-1307
  - name: "Xiaomeng Li"
    affiliation: "University of Connecticut"
  - name: "Timothy E. Moore"
    affiliation: "University of Connecticut"
    orcid_id: 0000-0002-9576-0517
output: distill::distill_article
---

@@

This paper shows how to use historical data on an attorney's use of peremptory strikes in past trials to estimate, in a current trial, how likely an attorney has used peremptory challenges because of prospective juror race or gender in violation of _Batson v. Kentucky_ (1986), and its progeny.

In the US, individuals selected for jury-service appear in court as scheduled and are questioned by the parties' attorneys and the trial judge.  During this "voir dire" process, a prospective juror, if not excused by the trial judge for cause, may still be dismissed (declared ineligible for a seat on the jury) if a party's attorney uses a peremptory challenge against them.

@@

By law, evidence supporting a _Batson_ challenge may include not only the circumstances of the jury selection itself, but also data on peremptory challenges in past cases. For example, in a _Batson_ challenge against a prosecutor in a criminal case, such historical data would be relevant to whether (1) to conclude that the defendant has a prima facie case; (2) to believe a prosecutor from that office who has offered a race-neutral explanation for striking, say, an African-American juror in that case; and (3) to ultimately accept or reject a Batson challenge in that case [e.g., Miller-El v. Dretke, 545 U.S. 231, 253 (2005); Riley v. Taylor, 277 F.3d 261, 280-81, 283 (3d Cir. 2001); Cochran v. Herring, 43 F.3d 1404, 1412 (11th Cir. 1995)]. The US Supreme Court recently identified evidence for a _Batson_ challenge as including the "relevant history of the State's peremptory strikes in past cases" [Flowers v. Mississippi, 139 S. Ct. 2228, 2243 (2019)].

Prior work on statistical methods for _Batson_ challenges, however, has largely focused on strike data from a single case [e.g. @Barrett2007; @Gastwirth2016; @Kadane2018; @Kadane2018a], perhaps because strike data from past cases is hard to collect. Jury-selection records are often not readily available and largely on paper, not electronic or machine-readable formats [@Grosso2017; @Wright2018], and otherwise require painstaking efforts to collect [e.g., @Craft2018]. As a result, even where there is, for any given individual attorney or prosecutor's office, available strike data on past trials, that data may be missing strike data from trials not at random [e.g., @Baldus2001, 132-134]. Possible reasons include limited access the relevant records (e.g., jury-selection records kept by a jury administrator, not in a case file) and the expense of obtaining copies. Some jurisdictions do not require recording relevant or detailed enough information about jurors who were or could have been struck, such as their race or gender. As a result, lawyers and researchers must resort to other data sources (e.g., voter registration rolls) to find that information. Finally, regardless of record-keeping policy, judges and court staff vary in how often they fully complete or maintain the required information.

When collecting strike data in multiple trials, prior studies have aggregated strikes, reported the observed difference in strike rates by race or gender, and tested for the probability of observing a non-zero difference in strike rates by race or gender, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference. For example, Eisenberg [-@Eisenberg2017, 334-339] used a 2-tail Fisher's exact test on the observed difference in peremptory strikes by race and gender in a non-random sample of thirty-five trials from 1997-2012 that resulted in death sentences (out of 63 such trials). @Baldus2001 similarly analyzed 317 venires (out of 461) in Philadelphia capital murder cases from which juries were selected between 1981 and 1997.

In contrast, @Gastwirth2013 criticized the use of a paired t-test to compare the differences in strike rates of Black and non-Black venire members in the aggregate. Rather, both they and @Gastwirth2014 argued for applying the Cochran–Mantel–Haenszel test by taking the difference of the sum of the observed numbers of Black venire members challenged in each trial and the corresponding sum of the expected numbers of Blacks challenged, and then comparing it to the standard deviation of the difference.

@@

Here, we extend a Bayesian approach for deteching illegal bias for _Batson_ challenges based solely on an attorney's pattern of peremptory challenges in present strike data and and that attorney's strikes in past cases.

In short, our workflow is this: First, specify a model of the peremptory-strike process that best approximates the peremptory-strike process in the jurisdiction of interest. Second, assign a prior distribution to the bias parameter $b$ that reflects the legal requirement that the _Batson_ challenger bear the burden of proof. Third, using the strike data in past trials generated by the same peremptory-strike process, estimate a posterior distribution for $b$.  Finally, using data on the target attorney's strikes in the present case, use that posterior distribution as the prior for $b$ and update.

In so doing, we make two main contributions. First, we use the power prior to control how much the strike history from past cases affects the posterior distribution of the bias parameter $b$. The power prior is a way to create informative prior distributions based on historical data so as to control how much historical data affects the posterior distribution of some parameter of interest. The key insight is to raise the likelihood function based on the historical data to a power parameter [@Ibrahim2000; @Ibrahim2015].

@@

Following @Kadane2018 and @Barrett2007, we model a struck-jury peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence.

In the "blind strike" variation, both parties simultaneously, rather than alternately, exercise their peremptory challenges on all the potential jurors.  This may result in both sides using a strike each on the same potential juror.

For any given case $i$ in which jury selection occurs, let $j$ denote a peremptory strike used, and let $\delta_{ij}$ denote whether or not a party used that strike on a person who belongs to a "cognizable class". If "race" is the bias type of interest, the cognizable class is racial minority jurors ($\delta_{ij} =1$, 0 for White jurors).  If "gender" is the bias of interest, the cognizable class is female jurors ($\delta_{ij} = 1$, 0 for male jurors). 
In turn, let $c_{ij}$ denote the number of cognizable class members subject to strike; and $m_{ij}$ denote the number of cognizable class non-members subject to strike, such that $c_{ij}+m_{ij}$ is the total number of jurors potentially subject to strike. If there is no bias, the probability is $\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking a cognizable class member, and $1-\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking someone who does not belong to that class.

By adding one parameter $w$, we can measure the bias by different values of $w$ by defining the probability of a cognizable class member being struck to be $\frac{wc_{ij}}{wc_{ij}+m_{ij}}$. To avoid making the weight of the non-cognizable class be the reciprocal of the weight of cognizable class, let $b = log(w)$. Accordingly, for any given value of the bias parameter $b$, the probability of strike of a member from either class, or $Pr(\delta_{ij})$, is shown in equation \@ref(eq:model1)

@@

We also generated historical strike data of three sizes (same, double, and triple the size of current data) as well as seven different values of the bias parameter for generating the historical data ( _historical_ $b = {-3,-2,-1,0,1,2,3}$). We let historical $b$ differ from current $b$ (our parameter of interest) to simulate noise associated with historical strike data that is incomplete not at random.

Alternatively, it lets us simulate a real difference in the bias parameter that depends on a feature that the current trial has but that _all_ the past trials in the historical data do not. To illustrate, assume a prosecutor with two true bias values: $b > 0$ if the criminal defendant is Black, $b = 0$ if not. If our historical data contains only strike data in cases with non-Black defendants, then that prosecutor's historical $b$ will differ from that prosecutor's bias in the current case with a Black defendant.

In any case, by using the power prior, we can control the weight that we put on historical data by modifying the weight parameter $a$. If $a = 1$, the historical strike data is equally weighted with the present data on strikes. If $a < 1$, the historical data matters proportionally less. For reasons discussed above, we may wish to assign $a < 1$ to evaluate how sensitive the posterior of the bias parameter is to the historical data on strikes. Accordingly, in the simulation study, we tried different values of the weight parameter $a = (0.1,0.3,0.5,0.7,1)$ to show how that parameter affected modeling accuracy.

@@

In turn, we set three credible interval level (80\%, 90\% and 95\%) and recorded the coverage rates, i.e., the proportion of times those intervals actually contained the true value of the (current) bias parameter $b$. For instance, a 90\% coverage rate means that in 900 of 1000 model fits, a given credible interval contained what we fixed as the value of current $b$. Put another way, if we set $b>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero among the 1000 model fits.
